The Castro AMR Simulation Code: Current and Future Developments

### 2019-10-02 astronum_2019_Castro/astronum-castro-2019.tex
fixed some minor typos

\castro\ has been applied to models of Type Ia supernovae,
core-collapse supernovae, pair-instability supernovae, exoplanet
atmospheres, and most recently X-ray bursts (see, e.g.,
\cite{castro-ccsne,castro-pairinstability,polin:2019,wdmergerI} for
some example science applications).  A common challenge in modeling
these events is the range of length and timescales involved.  To
capture length scales, \castro\ uses adaptive mesh refinement (AMR), through
the \amrex\ library~\cite{amrex_joss}.  This allows us to focus the
computational effort of regions where burning or the flow is
important.  To address the range of timescales involved in
astrophysical explosions, we have developed a low Mach number
hydrodynamics code, \maestroex~\cite{maestroex}, built on the same
framework as \castro, that can model the subsonic convection that
often precedes explosive events.  Both codes are open source and
freely available on
github\footnote{\url{https://github.com/AMReX-Astro/}}.

@@

The SDC algorithm used in \castro\ follows the ideas of
\cite{dutt:2000,minion:2003}, and uses low order explicit advection
and implicit reaction updates in a correction equation that when
applied iteratively achieves high-order time-accuracy.  In an operator
split implementation, reaction and burning proceed without knowing
about the other process.  In contrast, the update in the SDC
formulation explicitly couples the processes together.  In \castro\ we
consider reactions and hydrodynamics, and write our conservative
system as:
\begin{equation}
\Uc_t = \Advs{\Uc} + \Rbs{\Uc}
\end{equation}
where $\Uc$ is the vector of conserved quantities, $\Advs{\Uc}$ is the
advective term (the divergence of the hydrodynamic fluxes along with
hydrodynamic sources), and $\Rbs{\Uc}$ are the reactive source terms.
The update is done at specific temporal nodes, the number of location
of which are picked to give the desired temporal accuracy.  The update
from one time node $m$ to $m+1$ appears as:
    \begin{align}
      \label{eq:sdc:general}
      \avg{\Uc}^{{m+1},(k+1)} = \avg{\Uc}^{m,(k+1)}
            &+ \delta t_m \left [ \avg{\Advs{\Uc}}^{m,(k+1)} - \avg{\Advs{\Uc}}^{m,(k)} \right ] \nonumber \\
            &+ \delta t_m \left [ \avg{\Rbs{\Uc}}^{{m+1},(k+1)} - \avg{\Rbs{\Uc}}^{{m+1},(k)} \right ] \nonumber \\
            &+ \int_{t^m}^{t^{m+1}} dt  \left (\avg{\Advs{\Uc}}^{(k)} + \avg{\Rbs{\Uc}}^{(k)}\right)
   \end{align}
We use a finite-volume formalism, so $\avg{\Uc}$ is the spatial average of the
conserved state in a zone, $\avg{\Advs{\Uc}}^m$ is the average
advective term in a zone at time $t^m$, and $\avg{\Rbs{\Uc}}^{k+1}$ is
the average of the reactive source at time $t^{m+1}$.  The second
superscript on each term, $(k)$ or $(k+1)$, is the iteration counter.  This
update is an implicit equation for the new state,
$\avg{\Uc}^{{m+1},(k+1)}$.  The last term in the update is an integral
over the sources constructed from the previous iteration's values at
each time node.  Each iteration of the SDC method increases the
temporal order of accuracy by one, up to the order of accuracy with
which the integral is constructed.

\castro\ implements both a second-order method (using a trapezoid rule
for the integral) and fourth-order method in space and time (using a
Simpsons rule for the integral and the spatial reconstruction of
\cite{mccorquodalecolella}).  We demonstrated that \castro\ achieves
the expected convergence on a wide variety of problems in
\cite{castro:sdc}.  At the moment, the method is limited to single
levels, but work is underway to extend this methodology to AMR.  For
problems where burning is important and can dominate the computational
expense, we expect these new SDC method will become the preferred
integration technique in the future.

@@

The original approach to parallelism in \castro\ was MPI + OpenMP,
with scaling on manycore architectures achieve using a tiling approach
to OpenMP~\cite{tiling,astronum:2017}.  The general approach is to
have MPI distribute the AMR grids across nodes and have OpenMP threads
work on regions of each grid (tiles) by passing the box describing the
tile into the computational kernel.  More recently we've ported
\castro\ to GPUs using CUDA, using the same computational kernels as
the MPI + OpenMP version.  When run on GPUs, each CUDA thread handles
the update of a single zone, simply by passing that zone index into
the computational kernel.  This reuse keeps the code base
managable---we don't need separate kernels for each
architecture---while allowing us to take advantage of current and next
generation supercomputers.  All of the solvers needed to run our core
science problems run on GPUs: the main unsplit PPM hydrodynamics
scheme~\cite{ppm,millercolella:2002}, self-gravity via multigrid with
isolated boundary conditions, thermal diffusion, and nuclear
reactions.  Our approach is to put the data on GPUs and then run all
of the kernels on the GPUs, minimizing data movement.  This has given
us enormous speed-ups.  Figure~\ref{fig:gpu} shows weak scaling on the
OLCF Summit machine for a WD merger problem~\cite{wdmergerI}.  When
using CPUs, we use 42 IBM Power 9 cores per node, while when using
GPUs we use 6 Nvidia Volta GPUs per node.  We see that the code
performance is about $10\times$ higher per node when using GPUs, while
both show excellent weak scaling.

@@

With the new fourth-order accurate reactive hydro solver, we believe
we can capture the dynamics of the spreading flame accurately with one
less refinement level.  This will allow us to expand to larger domains
while keeping the memory demands reasonable.  Full science simulations
will begin once we port the SDC framework to AMR and generalize the
fourth order solver to axisymmetric geometries.  Proof-of-concept single
level runs are running now, but the lack of AMR makes the very
expensive.

The large increase in performance of the code when run on GPUs enables
us to do away with the flame boosting.  All of the physics solvers
needed for the XRB simulations run on the GPUs: the hydrodynamics,
explicit diffusion, realistic equation of state and conductivities,
and the 13-isotope He burning network we are using.  Preliminary
estimates show that we can expect a speed-up of atleast $5\times$ when
comparing the NERSC Cori Haswell nodes (using CPUs) to the OLCF Summit
nodes (using GPUs).  Work underway with further optimize the GPU code,
especially the reaction networks.

>>>>

\castro\ has been applied to models of Type Ia supernovae,
core-collapse supernovae, pair-instability supernovae, exoplanet
atmospheres, and most recently X-ray bursts (see, e.g.,
\cite{castro-ccsne,castro-pairinstability,polin:2019,wdmergerI} for
some example science applications).  A common challenge in modeling
these events is the range of length and timescales involved.  To
capture length scales, \castro\ uses adaptive mesh refinement (AMR), through
the \amrex\ library~\cite{amrex_joss}.  This allows us to focus the
computational effort on regions where burning or the flow is
important.  To address the range of timescales involved in
astrophysical explosions, we have developed a low Mach number
hydrodynamics code, \maestroex~\cite{maestroex}, built on the same
framework as \castro, that can model the subsonic convection that
often precedes explosive events.  Both codes are open source and
freely available on
github\footnote{\url{https://github.com/AMReX-Astro/}}.

@@

The SDC algorithm used in \castro\ follows the ideas of
\cite{dutt:2000,minion:2003}, and uses low order explicit advection
and implicit reaction updates in a correction equation that when
applied iteratively achieves high-order time-accuracy.  In an operator
split implementation, reaction and burning proceed without knowing
about the other process.  In contrast, the update in the SDC
formulation explicitly couples the processes together.  In \castro\ we
consider reactions and hydrodynamics, and write our conservative
system as:
\begin{equation}
\Uc_t = \Advs{\Uc} + \Rbs{\Uc}
\end{equation}
where $\Uc$ is the vector of conserved quantities, $\Advs{\Uc}$ is the
advective term (the divergence of the hydrodynamic fluxes along with
hydrodynamic sources), and $\Rbs{\Uc}$ are the reactive source terms.
The update is done at specific temporal nodes, the number of locations
of which are picked to give the desired temporal accuracy.  The update
from one time node $m$ to $m+1$ appears as:
    \begin{align}
      \label{eq:sdc:general}
      \avg{\Uc}^{{m+1},(k+1)} = \avg{\Uc}^{m,(k+1)}
            &+ \delta t_m \left [ \avg{\Advs{\Uc}}^{m,(k+1)} - \avg{\Advs{\Uc}}^{m,(k)} \right ] \nonumber \\
            &+ \delta t_m \left [ \avg{\Rbs{\Uc}}^{{m+1},(k+1)} - \avg{\Rbs{\Uc}}^{{m+1},(k)} \right ] \nonumber \\
            &+ \int_{t^m}^{t^{m+1}} dt  \left (\avg{\Advs{\Uc}}^{(k)} + \avg{\Rbs{\Uc}}^{(k)}\right)
   \end{align}
We use a finite-volume formalism, so $\avg{\Uc}$ is the spatial average of the
conserved state in a zone, $\avg{\Advs{\Uc}}^m$ is the average
advective term in a zone at time $t^m$, and $\avg{\Rbs{\Uc}}^{k+1}$ \MarginPar{I think this should be $\avg{\Rbs{\Uc}}^{m+1}$?}is
the average of the reactive source at time $t^{m+1}$.  The second
superscript on each term, $(k)$ or $(k+1)$, is the iteration counter.  This
update is an implicit equation for the new state,
$\avg{\Uc}^{{m+1},(k+1)}$.  The last term in the update is an integral
over the sources constructed from the previous iteration's values at
each time node.  Each iteration of the SDC method increases the
temporal order of accuracy by one, up to the order of accuracy with
which the integral is constructed.

\castro\ implements both a second-order method (using a trapezoid rule
for the integral) and fourth-order method in space and time (using a
Simpsons rule for the integral and the spatial reconstruction of
\cite{mccorquodalecolella}).  We demonstrated that \castro\ achieves
the expected convergence on a wide variety of problems in
\cite{castro:sdc}.  At the moment, the method is limited to single
levels, but work is underway to extend this methodology to AMR.  For
problems where burning is important and can dominate the computational
expense, we expect this new SDC method will become the preferred
integration technique in the future.

@@

The original approach to parallelism in \castro\ was MPI + OpenMP,
with scaling on manycore architectures achieved using a tiling approach
to OpenMP~\cite{tiling,astronum:2017}.  The general approach is to
have MPI distribute the AMR grids across nodes and have OpenMP threads
work on regions of each grid (tiles) by passing the box describing the
tile into the computational kernel.  More recently we've ported
\castro\ to GPUs using CUDA, using the same computational kernels as
the MPI + OpenMP version.  When run on GPUs, each CUDA thread handles
the update of a single zone, simply by passing that zone index into
the computational kernel.  This reuse keeps the code base
managable---we don't need separate kernels for each
architecture---while allowing us to take advantage of current and next
generation supercomputers.  All of the solvers needed to run our core
science problems run on GPUs: the main unsplit PPM hydrodynamics
scheme~\cite{ppm,millercolella:2002}, self-gravity via multigrid with
isolated boundary conditions, thermal diffusion, and nuclear
reactions.  Our approach is to put the data on GPUs and then run all
of the kernels on the GPUs, minimizing data movement.  This has given
us enormous speed-ups.  Figure~\ref{fig:gpu} shows weak scaling on the
OLCF Summit machine for a WD merger problem~\cite{wdmergerI}.  When
using CPUs, we use 42 IBM Power 9 cores per node, while when using
GPUs we use 6 Nvidia Volta GPUs per node.  We see that the code
performance is about $10\times$ higher per node when using GPUs, while
both show excellent weak scaling.

@@

With the new fourth-order accurate reactive hydro solver, we believe
we can capture the dynamics of the spreading flame accurately with one
less refinement level.  This will allow us to expand to larger domains
while keeping the memory demands reasonable.  Full science simulations
will begin once we port the SDC framework to AMR and generalize the
fourth order solver to axisymmetric geometries.  Proof-of-concept single
level runs are running now, but the lack of AMR makes them very
expensive.

The large increase in performance of the code when run on GPUs enables
us to do away with the flame boosting.  All of the physics solvers
needed for the XRB simulations run on the GPUs: the hydrodynamics,
explicit diffusion, realistic equation of state and conductivities,
and the 13-isotope He burning network we are using.  Preliminary
estimates show that we can expect a speed-up of at least $5\times$ when
comparing the NERSC Cori Haswell nodes (using CPUs) to the OLCF Summit
nodes (using GPUs).  Work is underway to further optimize the GPU code,
especially the reaction networks.
